---
title: Modern human alleles differentially regulate gene expression in brain tissues - implications for brain evolution
subtitle: Plots, statistics
authors: Alejandro Andirkó, Cedric Boeckx
date: 2020
contact: munoz.andirko@ub.edu
--- 


This document contains all the code necessary to plot and analyse the data from the "Modern human alleles differentially regulate gene expression in brain tissues: implications for brain evolution" manuscript. Please run the blocks from `materials.Rmd` before this one.

# R data and dependencies setup

Change working directory to whichever directory you are using to replicate the project:
```{r packages, message=FALSE}
library(tidyverse)
library(dplyr)
library(ggpubr)
library(cowplot)
library(reshape2)
library(knitr)
library(sanzo)
library(ggrepel)
library(gassocplot)
library(gwasglue)
library(coloc)
library(pwr) #power tests
library(wmwpow) #power test wilcoxon
library(EnvStats)
library(FSA)
library(outliers) 
library(statmod)
library(pheatmap)
library(usedist)
library(viridis)
library(chisq.posthoc.test)
library(ieugwasr)
library(S4Vectors)
library(gprofiler2)
library(biomaRt)
library(effectsize)
library(regioneR)
library(rtracklayer)
library(BSgenome.Hsapiens.UCSC.hg19.masked)
  
#change to your project directory
opts_knit$set(root.dir = "/media/sandor/c21e866e-0b3b-40be-a6ba-8cd9c93c15441/work/GTEX/now/data/")

#Imports a list of dfs with the HF, derived files
temp <- list.files(path = "eQTL/hsap/", pattern="*")
hsap_files <- lapply(paste0("eQTL/hsap/",temp),
                 read_table2, 
                 col_names = FALSE)

#Seed for reproducibility
set.seed(1)

#Clean names - used for plots mostly
cleannames <- function (listfiles){
  listfiles <- lapply(listfiles, str_remove_all, ".v8.*")
  listfiles <- lapply(listfiles, str_replace_all, "_", " ")
  unlist(listfiles)
}

#Splits in up and downregulation
divide <- function(x) {
  split(x,  x$X9 > 0 )
}

#Transform wide data on up-down regulation figure for stats
statsdf <- function(input){
  df <- input %>% 
  spread(direction, value)
  df <- as.data.frame(df) 
  rownames(df) <-  cleannames(temp)
  df <- df[,-1]
  return(df)
}

regular_clump <- function(input) {
  #Getting rsids and p-values for clumping function itself to work
  data <- NULL
  data <- lapply(input, '[', "X2")
  names(data) <- cleannames(temp)
  data <- melt(data)
  data$pval <- unlist(lapply(input, '[', "X13"))
  #Factors
  data <- mutate(data, L1 = factor(L1, ordered=T))
  colnames(data) <- c("rsid", "tissue", "pval")
  clumped <- ld_clump(dat = data, clump_r2 = 0.1, clump_kb = 1000000) #Uses EUR for reference, r2 = 0.001
  return(clumped)
}

#Similar to before, but with unique elements step
uniq_clump <- function(input) {
  data <- NULL
  data <- lapply(input, '[', "X2")
  temp <- cleannames(temp)
  names(data) <- stringr::str_replace_all(temp, "Brain ", "")
  data <- melt(data)
  data$pval <- unlist(lapply(input, '[', "X13"))
  #Factors
  data <- mutate(data, L1 = factor(L1, ordered=T))
  #Gets unique elements for each tissue
  uniq <- data %>%  
    full_join(data, by="X2") %>% 
    group_by(X2) %>% 
    filter(n() == 1)  
  uniq <- uniq[1:3]  
  colnames(uniq) <- c("rsid", "tissue", "pval")
  uniq_clumped <- ld_clump(dat = uniq, clump_r2 = 0.1, clump_kb = 1000000) #Uses EUR for reference, r2 = 0.001
  return(uniq_clumped)
}

#Get number of up and down-regulating variants ; store in all df
preplot_count <- function(dataset) {
  up <- NULL
  down <- NULL
  unfilt <- NULL
  for (i in 1:15){
    up <- append(up, length(dataset[[i]][["TRUE"]]))
    down <- append(down, length(dataset[[i]][["FALSE"]]))
    unfilt <- append(unfilt, length(dataset[[i]][["FALSE"]])+length(dataset[[i]][["TRUE"]]))
  }
  #all is a tidy df containing with the number of up and downregulating eQTL
  all$Up <- up
  all$Down <- down
  allp <- melt(all)
  return(allp)
}

# A specific fill plot that get's repeated over the code a couple times
plot_updown <- function(dataset, title) {
  ggplot() +
    theme_minimal() +
    ggtitle(title)+
    geom_bar(aes(y = value, x = tissue, fill = variable), 
             data = dataset, 
             stat= "identity",
             position="fill",
             alpha = 0.75)+
    labs(y = "Percentage", x = "", fill = "") +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = sanzo.duo("c106")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    geom_hline(yintercept=0.50, linetype="dashed") 
}

```
  
# Visualizations

## TSS 
```{r TSS plot}
sh <- 'awk \'{print $2, $4}\' eQTL/materials/GTEx_Analysis_v8_eQTL/* > TSS_unfilt'
system(sh, intern = TRUE)

sh <- 'awk \'{print $2, $4}\' eQTL/hsap/* > TSS_gtex'
system(sh, intern = TRUE)

alltss <- read_table2("TSS_unfilt") #Change to your path
gtextss <- read_table2("TSS_gtex", col_names = FALSE) #Change to your path

ggplot(alltss, aes(x=alltss$tss_distance)) + 
  geom_histogram(fill = viridis(79), bins = 80) +
  theme_minimal()+
  labs(x = "TSS distance", y = "Counts") +
  scale_x_continuous(name = "Counts", labels = scales::comma, breaks =c(-1000000, -500000, 0, 500000, 1000000), limits= c(-1000000, 1000000))

pdf("gtex_tss.pdf")
ggplot(gtextss, aes(x=gtextss$X2)) + 
  geom_histogram(fill = viridis(79), bins = 80) +
  theme_minimal()+
  labs(x = "TSS distance", y = "Counts") +
  scale_x_continuous(name = "Counts", labels = scales::comma, breaks =c(-1000000, -500000, 0, 500000, 1000000), limits= c(-1000000, 1000000))
dev.off()

sh <- 'rm TSS*'
system(sh, intern = TRUE)

```

## Heatmap with distance measures
```{r heatmap, message = FALSE}
#Unclumped since the point is to see how similar tissues are when only taking into account HF eQTL (regardless of eQTL specificity or other effects)

#Tidy data + give it long format
data <- NULL
data <- lapply(hsap_files, '[', "X9")
data <- melt(data)
data <- data %>% 
  group_by(L1) %>% 
  summarize(mean(value))
data <- data[,-1]
data <- as.data.frame(data)

names <- cleannames(temp)
names <- stringr::str_replace_all(names, "Brain ", "")
rownames(data) <- names

test <- dist(data, method = "euclidean")
test <- dist_setNames(test, names)
test <- as.matrix(test)
pdf("dist.pdf")
pheatmap(test, cluster_cols = TRUE, cluster_rows = TRUE, main = c("Cluster analysis of normalized effect size"))
dev.off()


```

## Clumped tissue-specific variants
```{r Clumpled unique tissues, message= FALSE}
#Apply function to selected tissues
uniq_clumped <- uniq_clump(hsap_files)
order <- forcats::fct_infreq(uniq_clumped$tissue)

pdf("unique_clumped.pdf")
ggplot(uniq_clumped, aes(x= forcats::fct_infreq(tissue))) +
  theme_minimal() +
  geom_bar(fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Tissue specific eQTL") + 
  scale_x_discrete(drop = FALSE) +
  labs(y = "Count", x = "Tissue") 
dev.off()


#Correlation test
GTExPortal <- read_csv("eQTL/GTExPortal.csv")
statsdf <- NULL
statsdf$samples <- GTExPortal[4]
statsdf$nrvars <- uniq_clumped %>% 
  group_by(tissue) %>% 
  summarize(n())

#Add adrenal and amygdala, which don't have unique rsids
statsdf$nrvars <- append(statsdf$nrvars[[2]], c(0,0), after = 0)
statsdf <- as.data.frame(statsdf)
rownames(statsdf) <- cleannames(temp)

#Testing for normality
shapiro.test(statsdf$X..RNASeq.Samples) #normal
shapiro.test(statsdf$nrvars) #non normal

#Correlation test
cor.test(statsdf$X..RNASeq.Samples, statsdf$nrvars, method = c("spearman"))
# Not significant
pwr.r.test(n = sum(statsdf$nrvars), r = -0-01, sig.level = 0.05)
#power = 1

#phewas analysis
#phew <- phewas(uniq_clumped$rsid)
```

## Fill plot with down and up-regulation proportions
```{r Fill clumped and unclumped, message = FALSE}
# Variant 1: Controlling for LD
clumped <- regular_clump(hsap_files)
# Note that clumped df contains duplicated rsids
# actually only 390 unique rsids, out of which ~ 200  are tissue-specific (as per uniq_clumped function)
clumped <- lapply(hsap_files, function(x) filter(x, X2 %in% clumped$rsid))
# Substituting df has to have same length as clumped before

divided_clumped <- lapply(clumped, divide)
all <- NULL
temp <- cleannames(temp)
temp <- stringr::str_replace_all(temp, "Brain ", "")
all$tissue <- temp
all <- as.data.frame(all)
divided_clumped <- lapply(divided_clumped[], function(x) lapply(x, '[[', 2))

pdf("figure3B_clumped.pdf")
divided_clumped <- preplot_count(divided_clumped)
two <- plot_updown(divided_clumped, "Directionality of regulation - clumped minor ancestral variants")
dev.off()

# Stats
print("Clumped data:")
colnames(divided_clumped)[2] <- c("direction")
stats_updown1 <- statsdf(divided_clumped)
chisq.test(stats_updown1,  correct = TRUE) # not significant
#chisq.posthoc.test(stats_updown1) # not needed

#Unclumped variant of the plot:
divided <- lapply(hsap_files, divide)
all <- NULL
temp <- cleannames(temp)
all$tissue <- temp
all <- as.data.frame(all)
divided <- lapply(divided[], function(x) lapply(x, '[[', 2))

pdf("figure3B_unclumped.pdf")
divided <- preplot_count(divided)
one <- plot_updown(divided, "Directionality of regulation - unclumped minor ancestral variants")
dev.off()

# Stats
colnames(divided)[2] <- c("direction")
stats_updown1 <- statsdf(divided)
print("Unclumped data:")
chisq.test(stats_updown1,  correct = TRUE) #significant

#Power test
samp <- c(sum(stats_updown1$Up)+sum(stats_updown1$Down))
phi <- chisq_to_cohens_w(143.16, samp, 15, 2)
phi # = w, but non numerical output of function
pwr.chisq.test(w = 0.13, N= samp, df = 14, sig.level=0.05) #Power = 1
#Post-hoc residuals analysis
chisq.posthoc.test(stats_updown1)

```

### Do ancestral major alleles also downregulate expression? (as opposed to ancestral minor alleles)
First, retrieving those eQTL:
```{sh Ancestral major alleles retrieval}
#Get a list of eQTL paired with file name; clean a bit the folder structure with sed
cd eQTL/materials
awk '{print $2, FILENAME, $7, $9}' GTEx_Analysis_v8_eQTL/* | sed 's/GTEx.*\///g ; s/\..*out//g' | tail -n +2 > list_eQTL.txt
#Since data from K&B (2019) already contains a catalog of which sites are different between humans and at least one archaic, a reverse grep of the list of eQTL sites through this database (NNaall.tsv) gives you a list of (in principle) non-derived eQTLs
# Then filter by MAF in these non Homo sapiens-specific variants
grep -Po 'rs\d*' NNaall.tsv | grep -vwf - list_eQTL.txt | awk '{if ($3<=0.10) print $0}' > nonderived_eQTL.txt
rm list_eQTL.txt

# We then have to filter our minor allele frequency to leave the  high frequency ancestral alleles
```

Now, plotting them:
```{r Fill major ancestral alleles}
anc_eQLT <- read_table2("eQTL/materials/nonderived_eQTL.txt", 
    col_names = FALSE)

anc_eQLT <- anc_eQLT %>% 
  mutate(direction = ifelse(X4>0,"Up","Down")) %>% 
  group_by(X2, direction) %>%
  summarize(value = n())

anc_eQLT$X2 <- cleannames(anc_eQLT$X2)
anc_eQLT$X2 <- stringr::str_replace_all(anc_eQLT$X2, "Brain ", "")

pdf("hf_ancestral_variants.pdf")
ggplot() +
  theme_minimal() +
  ggtitle("Directionality of regulation - HF ancestral variants")+
  geom_bar(aes(y = value, x = X2, fill = direction), 
           data = anc_eQLT, 
           stat= "identity",
           position="fill",
           alpha = 0.75)+
  labs(y = "Percentage", x = "", fill = "") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = sanzo.duo("c106")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept=0.50, linetype="dashed") 
dev.off()

# Are HF eQTL significantly skewed in up/down regulation?
stats_updown2 <- statsdf(anc_eQLT)
chisq.test(stats_updown2,  correct = TRUE) # Yes (< 2.2e-16)
#Power test
samp <- c(sum(stats_updown2$Up)+sum(stats_updown2$Down))
phi <- chisq_to_cohens_w(3653.7, samp, 15, 2)
phi # = w, but non numerical output of function
pwr.chisq.test(w = 0.05, N= samp, df = 14, sig.level=0.05) #Power = 1
#Post-hoc residual analysis
chisq.posthoc.test(stats_updown2) # Most of them

stats_onlydown <- NULL
stats_onlydown$ancmajor <- stats_updown2$Down
stats_onlydown$ancminor <- stats_updown1$Down
# Requires running chunk 4 before
# Currently corresponds to UNclumped data
stats_onlydown <- as.data.frame(stats_onlydown)
rownames(stats_onlydown) <- cleannames(temp)
# Are proportions in major vs minor downregulating ancestral eQTL significantly different? 
chisq.test(stats_onlydown,  correct = TRUE) # Yes (< 2.2e-16)
#Power test
samp <- c(sum(stats_onlydown$ancmajor)+sum(stats_onlydown$ancminor))
phi <- chisq_to_cohens_w(333.56, samp, 15, 2)
phi # = w, but non numerical output of function
pwr.chisq.test(w = 0.02, N= samp, df = 14, sig.level=0.05) #Power = 1
chisq.posthoc.test(stats_onlydown) 
```


## Sample size effect on hsap (unclumped)
```{r Sample size effect, message = FALSE}
divided <- lapply(hsap_files, divide)
#Store these in 'all' df
all <- NULL
temp <- cleannames(temp)
all$tissue <- temp
all <- as.data.frame(all)
divided <- lapply(divided[], function(x) lapply(x, '[[', 2))

unfilt <- NULL
for (i in 1:15){
  unfilt <- append(unfilt, length(divided[[i]][["FALSE"]])+length(divided[[i]][["TRUE"]]))
}

GTExPortal <- read_csv("eQTL/GTExPortal.csv")
unfiltnrs <- NULL

unfiltnrs$tissue <- cleannames(temp)
unfiltnrs$tissue <- stringr::str_replace_all(unfiltnrs$tissue, "Brain ", "")
unfiltnrs$value <- unfilt
unfiltnrs$sample <- GTExPortal[[4]]
unfiltnrs <- as.data.frame(unfiltnrs)

shapiro.test(unfiltnrs$sample) #normal distribution
shapiro.test(unfiltnrs$value) #slightly unnormal distribution (p-val = 0.02)

# No LD control violates independence assumption, so no tests here
pdf("figure3c.pdf")
ggplot(unfiltnrs, aes(x = sample, y= value)) +
  theme_minimal() +
  theme(legend.position="top") +
  labs(y= "Number of eQTL", x = "Sample size") +
  geom_smooth(alpha = 0.25) +
  geom_label_repel(aes(label = tissue), label.size = 0.05, box.padding = 0.7) +
  geom_point(size = 3.5) 
dev.off()

```

## Sample size effect figure, clumped:
```{r 3C clumped, messages = FALSE}
# Controlling for LD
clumped <- regular_clump(hsap_files)
clumped <- lapply(hsap_files, function(x) filter(x, X2 %in% clumped$rsid))
divided_clumped <- lapply(clumped, divide)
divided_clumped <- lapply(divided_clumped[], function(x) lapply(x, '[[', 2))

unfilt <- NULL
for (i in 1:15){
  unfilt <- append(unfilt, length(divided_clumped[[i]][["FALSE"]])+length(divided_clumped[[i]][["TRUE"]]))
}

GTExPortal <- read_csv("eQTL/GTExPortal.csv")
unfiltnrs <- NULL
unfiltnrs$tissue <- cleannames(temp)
unfiltnrs$tissue <- stringr::str_replace_all(unfiltnrs$tissue, "Brain ", "")
unfiltnrs$value <- unfilt
unfiltnrs$sample <- GTExPortal[[4]]
unfiltnrs <- as.data.frame(unfiltnrs)

pdf("Figure_3C_clumped.pdf")
ggplot(unfiltnrs, aes(x = sample, y= value)) +
  theme_minimal() +
  theme(legend.position="top") +
  labs(y= "Number of eQTL", x = "Sample size") +
  geom_smooth(alpha = 0.25) +
  geom_label_repel(aes(label = tissue), label.size = 0.05, box.padding = 0.7) +
  geom_point(size = 3.5) +
  stat_cor(method = "spearman", label.x = 120, label.y = 200,color="black") 
dev.off()

# Spearman: sample ~ value correlation 
cor.test(unfiltnrs$sample, unfiltnrs$value, method = "spearman", conf.level = 0.95)
# Significant, as seen in plot
# p-value = 1.717e-05
pwr.r.test(n = 211, r = 0.7363, sig.level = 0.05) #power 1 
```


# Other Analysis

## Variant functional consequences
Retrieve variants to query Biomart
```{sh Variants for Biomart}
cd eQTL/materials
awk '{print $2}' GTEx_Analysis_v8_eQTL/* | sed 's/,/\n/g' | sort | uniq > snps
```


```{r BiomaRt}
snps <- read_table("eQTL/materials/snps", col_names = "snps")

ensembl = useEnsembl(biomart="snp", dataset = "hsapiens_snp", GRCh = 37)

result <- getBM(attributes=c('consequence_type_tv', 'refsnp_id'),
  filters = 'snp_filter',
  values = snps$snps,
  mart = ensembl)
write.csv(result, "eQTL/materials/result.csv")
```

BiomaRt is not very good with large queries, so if you find that the package crashes at any point of the batch submision, or it takes too long, another option would be to run the following code chunk. Note that the subsequent code in the section is based on the results given by the biomaRt package, and, thought the data itself shouldn't vary, we warn that there might be differences given the way JSON queries are returned:

```{r BiomaRt REST}
library(httr)
library(jsonlite)
library(xml2)

server <- "https://grch37.rest.ensembl.org"
prefix <- "/variation/human/"

getconsequence <- function(rsid){
  query <- paste0(prefix, rsid, "?")
   r <- GET(paste0(server,query), content_type("application/json"))
   stop_for_status(r)
   return(data.frame(t(sapply(content(r),c))))
}

l <- NULL
for (rsid in snps$snps){
  result <- getconsequence(rsid)
  l[[rsid]] <- result
}
```

From the queried snps, get only those that are NOT in HF
```{sh Get noHF snps}
cd eQTL/materials

grep -Po "rs\d*" result.csv | grep -vwf hfsapiens.freqfilt.tsv -  > result_noHF.csv
rm result.csv
```


```{r BiomaRt non HF}
result_noHF <- read_table2("eQTL/materials/result_noHF.csv",
                           col_types = cols(POS = col_character()))
#NA ommited

result_noHF.noNA <- na.omit(result_noHF$dbSNP) 

dist <- result_noHF.noNA %>% 
  group_by(consequence) %>% 
  dplyr::summarize(count = n())

dist$consequence <- str_replace_all(dist$consequence, "_", " ")
dist$consequence <- str_to_title(dist$consequence)
dist$consequence <-factor(dist$consequence, levels = dist$consequence[order(dist$count)])
dist$count <- as.factor(dist$count)

ggplot(dist, aes(x = consequence, y = count)) +
  theme_minimal() +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_colour_viridis_d() +
  coord_flip() +
  labs(x = "Variant consequence", y = "Nr. of eQTL loci") +
  theme(axis.text.x = element_text(angle = 90)) 

#Chi-square test of independence
#Data as in dist (all eQTL, without the high freq ones & (HF, Homo Sapiens specific eQTL variants only)
#More categories in the first one than the second: 
#Only the second set of categories (present in HF variants and non-HF variants) is subjected to testing here 
dist_chi <- data.frame("Features" = c("intron", 
                                      "non coding transcript",
                                      "NMD transcript",
                                      "non coding trans exon var",
                                      "3prime UTR",
                                      "5prime UTR",
                                      "Missense",
                                      "Synonymous",
                                      "Splice reg",
                                      "Stop lost"),
                       "All_eQTL_minusHF" = c(998774,576017,282909,59840,37243, 12846,12334,11814,3938,417), 
                       "HS_HF_eQTL" = c(3453,2190,843,277,131,98,56,47,16,1))
chisq.test(dist_chi[2:3], correct = TRUE)

#Power test
samp <- c(sum(dist_chi[2])+sum(dist_chi[3]))
phi <- chisq_to_cohens_w(121.72, samp, 10, 2)
phi # = w, but non numerical output of function
pwr.chisq.test(w = 7.79e-03, N= samp, df = 9, sig.level=0.05) #Power = 1

#Post-hoc analysis
chisq.posthoc.test(dist_chi[2:3], method = "bonferroni")
```


## Random sampling analysis
Is the number of tissue-specific eQTL variants independent of sample size bias?

```{r Random sampling}
data <- NULL
data <- lapply(hsap_files, '[', "X2")
temp <- cleannames(temp)
names(data) <- temp
data <- melt(data)
data <- unique(data)

#Factors
data <- mutate(data, L1 = factor(L1, ordered=T))

#Gets unique elements for each tissue
uniq <- data %>%  
  full_join(data, by="X2") %>% 
  group_by(X2) %>% 
  filter(n() == 1)  

uniq <- uniq_clump(hsap_files)
#GTExPortal <- read_csv("eQTL/GTExPortal.csv")

#Collecting random samples
sampling <- function (n) {
  samp <- sample(n[[2]], 100, replace = FALSE)
  sum(samp %in% uniq[[1]])
}

sampled <- lapply(hsap_files, sampling)
sampled <- unlist(sampled)
sampled <- as.data.frame(sampled)
sampled$nonuniq <- 100-sampled$sampled
rownames(sampled) <- cleannames(temp)
sampled <- sampled %>% 
  relocate(sampled, .after = nonuniq)

chisq.test(sampled)
#Power test
samp <- c(sum(sampled[1])+sum(sampled[2]))
phi <- chisq_to_cohens_w(129.35, samp, 15, 2)
phi # = w, but non numerical output of function
pwr.chisq.test(w = 0.29, N= samp, df = 14, sig.level=0.05) #Power = 1
chisq.posthoc.test(sampled)
```


## Total ammount of variance
Is the number of HF eQTL dependent on total amount of variance for all tissues?

```{r Total variance}
#Setting up object for table
statsdf <- NULL

sh <- 'wc -l eQTL/hsap/* | awk \'{print $1}\' | head -n -1'
statsdf$hf_hsap <- system(sh, intern = TRUE)

sh <- 'wc -l eQTL/materials/GTEx_Analysis_v8_eQTL/* | awk \'{print $1}\' | head -n -1'
statsdf$total <- system(sh, intern = TRUE)

statsdf <- lapply(statsdf, as.numeric)
#Check is same order of files
hs_files <- list.files(path = "eQTL/hsap/", pattern="*")
total_files <- list.files(path = "eQTL/materials/GTEx_Analysis_v8_eQTL/", pattern="*")
hs_files <- cleannames(hs_files)
total_files <- cleannames(total_files)
hs_files == total_files 
#Should give TRUE


# Minus successes (=hf homo sapiens-specific variants)
statsdf$total <- as.numeric(statsdf$total)-as.numeric(statsdf$hf_hsap)
statsdf$total

# Add names of tissues 
statsdf <- as.data.frame(statsdf)
rownames(statsdf) <- hs_files

chisq.test(statsdf, correct=TRUE) 
#Significant  
#Power test
samp <- c(sum(statsdf[1])+sum(statsdf[2]))
phi <- chisq_to_cohens_w(577.89, samp, 15, 2)
phi # = w, but non numerical output of function
pwr.chisq.test(w = 6.59e-03, N= samp, df = 14, sig.level=0.05) #Power = 1

chisq.posthoc.test(statsdf)
#Significant for about half the tissues, so not only one driving the effect 
```

## H test

Selection test that doesn't require an archaic sample:
```{sh Fay and Wu s H (shell)}
cd eQTL
mkdir FayWuH
cd FayWuH
wget http://hsb.upf.edu/hsb_data/positive_selection_NAR2013/FayWuH_CEU.whole_genome.pvalues.gz
wget http://hsb.upf.edu/hsb_data/positive_selection_NAR2013/FayWuH_CHB.whole_genome.pvalues.gz
wget http://hsb.upf.edu/hsb_data/positive_selection_NAR2013/FayWuH_YRI.whole_genome.pvalues.gz
```

```{r Fay and Wu s H (R), message=FALSE}
CEU <- read_table2("eQTL/FayWuH/FayWuH_CEU.whole_genome.pvalues.gz")
CHB <- read_table2("eQTL/FayWuH/FayWuH_CHB.whole_genome.pvalues.gz")
YRI <- read_table2("eQTL/FayWuH/FayWuH_YRI.whole_genome.pvalues.gz")

# P-values are in -log10 scale
# That means a higher than 2 score in the pvalue column = a < 0.01 p-value
filterpval <- function(popfile){ 
  popfile <- popfile %>% 
    filter(pvalue > 2) %>% 
    select(chromosome, start, end, score, pvalue)
  as.data.frame(popfile)
}

CEU <- filterpval(CEU)
CHB <- filterpval(CHB)
YRI <- filterpval(YRI)


#Prepare eQTL bed-like dataframe
data <- lapply(hsap_files, '[', "X1")
temp <- cleannames(temp)
names(data) <- temp
data <- melt(data)
data <- data %>% 
  separate(X1, sep = "_", into = c("chr","start","alleleA","alleleB","erase")) %>% 
  select(chr, start, L1)

data$end <- data$start
data <- data[, c(1,2,4,3)]

CEUeqtl <- overlapRegions(CEU, data)
CHBeqtl <- overlapRegions(CHB, data)
YRIeqtl <- overlapRegions(YRI, data)
#No overlaps
```

## MAF of SNPs overlapping and not overlapping w/ eQTL

```{sh MAF eQTL}
#Get a list of eQTL with MAF (already in file)
cd eQTL/materials
rm list_eQTL.txt #list_eQTL.txt already exists in folder, but with a 10% filter for - ensure it's not used for previous chunks
awk '{print $2, $7}' GTEx_Analysis_v8_eQTL/* | grep -v "maf" | sort | uniq > list_eQTL.txt
head list_eQTL.txt

#Pre not-an-eQTL filtering ; that'll be done in the R chunk
awk '$19 ~ /rs[:digit:]*/ {print $19, $2}' NNaall.tsv | sed 's/\,.*//g' > list_NNaallrsids.txt

```

``` {r MAF eQTL}
list_eQTL <- read_table2("eQTL/materials/list_eQTL.txt", 
    col_names = FALSE)

NNaall_prefilt <- read_table2("eQTL/materials/list_NNaallrsids.txt", 
    col_names = FALSE)

no_eQTL <- NNaall_prefilt %>% 
  filter(!(X1 %in% eQTL$X1))
yes_eQTL <- NNaall_prefilt %>% 
  filter(X1 %in% eQTL$X1)

no_eQTL <- na.omit(no_eQTL)
yes_eQTL <- na.omit(yes_eQTL)

#around 10% mean difference
mean(no_eQTL$X2) #0.3228581
mean(yes_eQTL$X2) #0.429301

#Controlling for size
set.seed(1)
test <- sample_n(no_eQTL, length(yes_eQTL$X1))
```

## GO analysis
With clumped data to avoid noise.

```{r GO}
clumped <- regular_clump(hsap_files)


results_clumpedGO <- NULL
results_clumpedGO <- gost(query = clumped$rsid,
     organism = "hsapiens", ordered_query = FALSE,
     multi_query = FALSE, significant = TRUE, exclude_iea = FALSE,
     measure_underrepresentation = FALSE, evcodes = FALSE,
     user_threshold = 0.05, correction_method = "g_SCS",
     domain_scope = "annotated", custom_bg = NULL,
     numeric_ns = "", sources = NULL, as_short_link = FALSE)


data <- NULL
query <- data.table::rbindlist(lapply(hsap_files, `[`, 2))
results_unclumpedGO <- NULL
results_unclumpedGO <- gost(query = query$X2,
     organism = "hsapiens", ordered_query = FALSE,
     multi_query = FALSE, significant = TRUE, exclude_iea = FALSE,
     measure_underrepresentation = FALSE, evcodes = FALSE,
     user_threshold = 0.05, correction_method = "g_SCS",
     domain_scope = "annotated", custom_bg = NULL,
     numeric_ns = "", sources = NULL, as_short_link = FALSE)

#How about cerebellum unique SNPs in particular?
uniq_clumped <- uniq_clump(hsap_files)
test <- uniq_clumped %>% 
  group_by(tissue) %>%  
  filter(tissue == "Brain Cerebellum") %>% 
  dplyr::select(rsid)

result <- gsnpense(query = test[[2]])


results_clumpedGO <- gost(query = result$ensgs,
                          organism = "hsapiens", ordered_query = FALSE,
                          multi_query = FALSE, significant = TRUE, exclude_iea = FALSE,
                          measure_underrepresentation = FALSE, evcodes = FALSE,
                          user_threshold = 0.05, correction_method = "g_SCS",
                          domain_scope = "annotated", custom_bg = NULL,
                          numeric_ns = "", sources = NULL, as_short_link = FALSE)



```


## Colocalization analysis
Change the annotated line to test for other traits. The following IDs were tested:
ieu-a-806 # autism
prot-a-498 # cerebral dopamine
ubm-a-164 # IDP T1 FAST ROIs R cerebellum X
ubm-a-150 #IDP T1 FAST ROIs R cerebellum
ieu-b-32 # lymphocyte cell count
ukb-b-14699 # Alzheimer's disease
ubm-a-2659 , ubm-a-2658, ubm-a-2676, ubm-a-2677 # Cerebellar volume

```{r Colocalization}
#Modify eQTL dataframe
clumped <- uniq_clump(hsap_files) 
clumped <- lapply(hsap_files, function(x) filter(x, X2 %in% clumped$rsid))
clumped <- lapply(clumped, rbind)
clumped <- bind_rows(clumped)

gwas <- phewas(clumped$X2, pval = 0.0001)

total <- separate(clumped, col = X1, into= c("chr","pos", "ref", "alt", "ver"), sep = "[_]")
total$chr <- str_replace_all(total$chr, "chr", "")
colnames(total)[6] <- c("rsid")


colocal_quick <- function(instrum){
# Extract results GWAS
  top <- extract_instruments(outcomes=instrum, clump = FALSE, p1 = 0.0001)
  colnames(top) <- stringr::str_replace_all(colnames(top), ".exposure", "")
  colnames(top)[8] <- c("rsid")
  #Example with schizophrenia (phewas)
  #top <- ieugwasr::tophits('ieu-b-42', clump = 0, pval = 5e-04) %>% dplyr::arrange(p)
  #Merge!
  input <- merge(top, total, by="rsid", all=FALSE, suffixes=c("eqtl","gwas"))
  return(input)
}

test <- colocal_quick('prot-a-2235') #substitute ID for relevant instrument

my.res <- coloc.abf(dataset1=list(beta=input$beta, varbeta=input$se, N=nrow(top),type="quant"),
                    dataset2=list(beta=input$X7, varbeta=input$X7, N=nrow(total),type="quant"),
                    MAF=input$eaf)

# Position of top variant
chrpos <- paste0(input$chr[1], ":", input$position[1] - 90000, "-", input$position[1] + 90000)

temp <- coloc_to_gassocplot(my.res)
gassocplot::stack_assoc_plot(temp$markers, temp$z, temp$corr, traits=temp$traits)

```

## Regions of evolutionary significance
Gene name retrieval:

```{sh Gene names selection}
cat eQTL/hsap/*.rs > all.rs

awk '{print $1, $2, $3}' all.rs | sed 's/_/ /g' | awk 'NR>1 {print $1, $2-1, $2, $6, $7}' | sed 's/\s/\t/g' > coords.bed
# $2-1 due to the reasons explained here: https://bedtools.readthedocs.io/en/latest/content/overview.html#bed-starts-are-zero-based-and-bed-ends-are-one-based  

#Download the positive selections tables formatter for your convenience
#You can retrieve them if you'd prefer from the supplementary information of the two studies
wget --no-check-certificate 'https://docs.google.com/uc?export=download&id=1LBr_oepH8yhveTuz1w7rd3Jznnqp-iw1' -O racimo_coords.txt
wget --no-check-certificate 'https://docs.google.com/uc?export=download&id=1IY2NCdCsgW_Mj48ygTVzOaXw7ag-CN0P' -O pey_coords.txt

bedtools intersect -a coords.bed -b pey_coords.txt > ps_pey
bedtools intersect -a coords.bed -b racimo_coords.txt > ps_rac
rm all.rs racimo_coords.txt pey_coords.txt coords.bed
```

```{r Gene names selection}
library(gprofiler2)
library(readr)

#Peyregne
table_peyregne <- read_delim("ps_pey", "\t", 
                     escape_double = FALSE, col_names = FALSE, 
                     trim_ws = TRUE)

#Residing gene names retrieval
table_peyregne$X5 <- gsub("[.]\\d*$", "", table_peyregne$X5)
result_query <- gconvert(query = table_peyregne$X5,
         organism = "hsapiens",
         target = "ENSG",
         filter_na = TRUE)
table_peyregne$eGene <- c(result_query$name)
    
result_query <- gconvert(query = table_peyregne$X4,
                    organism = "hsapiens",
                    target = "ENSG",
                    filter_na = FALSE)
table_peyregne$target <- c(result_query$name)


#Racimo
table_racimo <- read_delim("ps_rac", "\t", 
                     escape_double = FALSE, col_names = FALSE, 
                     trim_ws = TRUE)
table_racimo$X5 <- gsub("[.]\\d*$", "", table_racimo$X5)
result_query <- gconvert(query = table_racimo$X5,
                    organism = "hsapiens",
                    target = "ENSG",
                    filter_na = TRUE)
table_racimo$eGene <- c(result_query$name)

result_query <- gconvert(query = table_racimo$X4,
         organism = "hsapiens",
         target = "ENSG",
         filter_na = TRUE)

table_racimo$target  <- ifelse(table_racimo$X4 %in% result_query$input, 
       result_query$name, 
       NA)
```


Now, as for the permutation tests:

```{sh Permutation tests}
cat eQTL/hsap/*.rs | awk '{print $1, $2, $9}' | sed 's/.*chr/hs/g' | sed 's/\_/ /g' | awk '{print $1, $2, $2, $7}' | sed -e 's/\s/\t/g' | awk '{print $1, $2, $3}' | sed 's/hs/chr/g' > snppermut.txt

wget --no-check-certificate 'https://docs.google.com/uc?export=download&id=1LBr_oepH8yhveTuz1w7rd3Jznnqp-iw1' -O racimo_coords.txt
wget --no-check-certificate 'https://docs.google.com/uc?export=download&id=1IY2NCdCsgW_Mj48ygTVzOaXw7ag-CN0P' -O pey_coords.txt

sed 's/hs/chr/g' racimo_coords.txt > racimo_perm
sed 's/hs/chr/g' pey_coords.txt > pey_perm
rm pey_coords.txt racimo_coords.txt
```

```{r Permutation tests}
pey <- read_table2("pey_perm", col_names = FALSE)
pey <- as.data.frame(pey)
pey <- toGRanges(pey)

racimo <- read_table2("racimo_perm", col_names = FALSE)
racimo <- as.data.frame(racimo)
racimo <- toGRanges(racimo)

data <- read_table2("snppermut.txt")
data <- as.data.frame(data)
data <- toGRanges(data)

#S4Vectors has deprecated a function needed for regioneR: make sure to use version 0.29.1 of S4Vectors
#remotes::install_github("Bioconductor/S4Vectors@2f2e26d8a1b58546d2e05672e28f41aba2ee0258")

pt1 <- permTest(A=data, B=pey, ntimes=1000,
                randomize.function=circularRandomizeRegions,
                evaluate.function=numOverlaps, mc.set.seed=TRUE, force.parallel=TRUE
)
pt2 <- permTest(A=data, B=racimo, ntimes=1000,
                randomize.function=circularRandomizeRegions,
                evaluate.function=numOverlaps, mc.set.seed=TRUE, force.parallel=TRUE)

pt1
#expected: mean(pt1[["numOverlaps"]][["permuted"]])
pt2
#expected: mean(pt2[["numOverlaps"]][["permuted"]])

pdf("peyregne_perm.pdf")
plot(pt1, plotType="Tailed")
dev.off()

pdf("racimo_perm.pdf")
plot(pt2, plotType="Tailed")
dev.off()
```

Per tissue testing: are some tissues more prone to be represented in these positive selection regions?
```{r Overlap positive selection}
pey <- read_table2("pey_perm", col_names = FALSE)
pey <- as.data.frame(pey)
pey <- toGRanges(pey)

racimo <- read_table2("racimo_perm", col_names = FALSE)
racimo <- as.data.frame(racimo)
racimo <- toGRanges(racimo)

data <- read_table2("snppermut.txt")
data <- as.data.frame(data)
data <- toGRanges(data)

listhf <- hsap_files %>% 
  melt(.) %>% 
  dplyr::select(X1, L1) 

#Renames tissues
names <- cleannames(temp)
renametiss <- function(x, y) {
  str_replace_all(x, paste0("^",y,"$"), names[y])
}
# For loop to iterate over L1 list index numbers
for (n in c(1:15)) {
  listhf$L1 <- renametiss(listhf$L1,n)
}

listhf <- separate(listhf, col = X1, into= c("chr","pos", "ref", "alt", "ver"), sep = "[_]")
#Remove duplicated columns - not taking into account multiple genes affected by one eQTL
listhf <- unique(listhf)

#Overlaps, and table with more common tissues in tiss_pey
overlap_pey <- overlapRegions(data,pey)
tiss_pey <- listhf %>% 
  filter((listhf$chr %in% overlap_pey$chr) & (listhf$pos %in% overlap_pey$startA))
tiss_pey <- melt(table(tiss_pey$L1))

overlap_rac <- overlapRegions(data,racimo)
tiss_rac <- listhf %>% 
  filter((listhf$chr %in% overlap_rac$chr) & (listhf$pos %in% overlap_rac$startA))
tiss_rac <- melt(table(tiss_rac$L1))

#dataframe to plot
plot <- NULL
plot$Peyregne <- tiss_pey$value
plot$Racimo <- tiss_rac$value
plot <- as.data.frame(plot)
names <- stringr::str_replace_all(names, "Brain ", "")
plot$tissues <- names
plotm <- melt(plot)

pdf("eQTLpossel.pdf")
ggplot(plotm, aes(x=tissues, y=value, color=variable)) + 
  theme_minimal() +
  geom_point(size=2) +
  ggtitle("eQTL in regions of positive selection") +
  labs(y = "Count", x = "Tissues", color = "Study") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dev.off()

pdf("eQTLpossel2.pdf")
plotm <- plotm %>% mutate(tissues = factor(tissues),  tissues = factor(tissues, levels = rev(levels(tissues))))
one <- ggplot(plotm,aes(x= tissues,y=value, fill = variable))+
		theme_minimal() +
		geom_bar(stat="identity", position = "dodge") +
		labs(title="Overlapping eQTL",
		 x = "Count",
	 	y = "Structures", 
	 	fill = "") +
		scale_fill_discrete(labels=c("Peyregné","Racimo"))+ # check this
    coord_flip()
dev.off()

plot

#Statistics: 
#First, controlling for the difference in window sizes between the two studies
#Import again
pey <- read_table2("pey_perm", col_names = FALSE)
pey <- as.data.frame(pey)
racimo <- read_table2("racimo_perm", col_names = FALSE)
racimo <- as.data.frame(racimo)

proportion <- sum(racimo$X3-racimo$X2)/sum(pey$X3-pey$X2)
plot$Racimo <- plot$Racimo/proportion

test <- wilcox.test(plot$Peyregne,
            plot$Racimo,
            paired=TRUE) #6.104e-05
test

#Power test for wilcoxon
shiehpow(sum(plot$Peyregne), sum(plot$Racimo), p=6.104e-05, dist = "norm") 
# shieh power:1

#Is cerebellum eQTL number not variable in the two studies compared to the rest of tissues?
diff <- plot$Peyregne-plot$Racimo
plot$diff <- diff
plot(plot$diff)

test <- dunnTest(value ~ tissues, data=plotm, method="bh")
test 

#Are selection tests coinciding in particular eQTL?
overlap_pey <- overlapRegions(data,pey)
tiss_pey <- listhf %>% 
  filter((listhf$chr %in% overlap_pey$chr) & (listhf$pos %in% overlap_pey$startA))

overlap_rac <- overlapRegions(data,racimo)
tiss_rac <- listhf %>% 
  filter((listhf$chr %in% overlap_rac$chr) & (listhf$pos %in% overlap_rac$startA))

bothstudies <- tiss_rac %>% 
  filter((tiss_rac$chr %in% tiss_pey$chr) & (tiss_rac$pos %in% tiss_pey$pos))


```
